---
title: "R Basics Course"
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_notebook:
    theme: spacelab
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
---

#Introduction
This course is an introduction into using R for data analysis. It contains a series of demonstrations on how to use basic functions and operators in R.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(dslabs)
library(tidyverse)
library(dplyr)
data(murders)
```


##Assorted Useful Functions

* `ls()` show all objects in workspace
* `args` provides arguments for a function
* `data()` lists all datasets provided in R
* `exp` and `log` exponentiate and take logs, and default to the natural
* `class` gives object type
* `str` gives details of an object (structure of dataset)
* `names` gives columns names of object
* `length` gives # of entries in a vector
* `levels` gives factors in column
* `identical` returns a logical: whether the two supplied objects are identical
* `table` returns the frequency of unique elements in a vector

* A useful note: using [[""]] rather than $ returns a data.frame rather than a vector

##Vectors

*Types of Vectors*

Vectors can hold different elements including numbers, strings, or named numbers.
```{r}
codes <- c(380,124, 818)
country <- c("italy", "canada", "egypt")
name_codes <- c(italy = 380, canada = 124, egypt = 818)
```

The function names() can be used to assign names to entries in a vector.
```{r}
names(codes) = country
codes
```

*Vector Coercion*

* Creating a vector with a string in it will lead R to coerce everything else to a string
    + one can force coercion with `as.numeric()`, `as.character()`, and `as.factor()`
    + if coercion fails, R returns an `NA`

*The Integer Class*

* A distinct class from numeric is the integer, can be created by adding "L" after a number
    + useful, because it takes up less space than a number

*Sorting and Finding Elements*
Vectors can be sorted in various ways.

* `sort` orders the column
* `order` produces an index which, when applied to the vector, would sort it
    + e.g. writing `index = order(x)` followed by`x[index]` returns x in ascending         order
    + one column can be ordered by the order of another using its index
* `max` returns the largest element
* `which.max` returns which element is the max
* `which.mix` is the same as above for the min
* `rank` returns the rank of each number in a vector (from lowest to highest, -x gives     highest to lowest)


##Data Frames

Data frames are a more versatile type of matrix that can hold multiple types of elements. 

```{r}
sample_dat <- data.frame(NameOfColumn1 = "VectorObject1", NameofColumn2 = "VectorObject2")
```

Some notes:
* arithmetic operations occur element-wise
* indexing refers to ordering based on a certain criterion
* a useful way to create a new column is to create a var from manipulation of columns
    + e.g. col1/col2 = col3; df$col3 = col3
* `stringsAsFactors` within the `data.frame` function can be used to prevent it from turning characters into factors

#Indexing Functions

* one can create an index by defining a var e.g. "index" as a logical
    + one can thaen subset based on the logical with data[index]
* `sum` coerces TRUE to numeric i.e. 1, so sum() of logical gives total TRUE
* To satisfy multiple conditions, create a variable for each, and then use
    the and operator "&" to create an index that contains each
* `which` gives entries for which something is true
* `match` give it a vector of names and a column, and it returns the row # where they    appear
* `%in%` checks to see if the elements of a first vector are in a second vector

An exmple using `which()` and `%in%` in concert to identify which components of one
vector are present in another:
```{r}
abbs <- c("MA", "ME", "MI", "MO", "MU")
ind <- which(!abbs%in%murders$abb)
abbs[ind]
```


##Dplyr Package

The `dplyr` package contains useful packages for manipulating datasets.

    install.packages("dplyr")
    library(dplyr)

*Mutate*

Create new columns with `mutate`.
```{r}
murders <- mutate(murders, rate= total/population*100000)
```

Note: total and population are not defined; mutate knows to look to column names

*Filter*

Select rows based on specified criteria with `filter`.

```{r}
filter(murders, rate < 0.71) %>% head()
```

*Select*

Subset the columns of a data frame with `select`.

```{r}
new_table <- select(murders, state, region, rate)
```

*The Pipe*

The pipe, part of the tidyverse, is a useful way to apply multiple operations to the same dataset without creating intermediate variables.

Rather than,
```{r}
new_table <- select(murders,state, region,rate)
filter(new_table, rate < 0.71) %>% head()
```

We can write:

```{r}
murders %>% select(state, region, rate) %>% filter(rate > 0.71) %>% head()
```

##Basic Plots


*Basic Plotting*

Base R provides the `plot` function for rendering scatterplots, 

```{r}
population_in_millions <- murders$population/10^6
total_gun_murders <- murders$total
plot(population_in_millions, total_gun_murders)
```

as well as the `hist` function for creating histograms:

```{r}
hist(murders$rate)
```

Boxplots are generated with `boxplot`:

```{r}
boxplot(rate~region, data = murders, main = "Distributions of Murder Rates by Region")
```

##Basic Programming in R

*If Else Statemenets*

A sample of an if-else statement:

```{r}
a <- 2
if(a!=0){
  print(1/a)
}else{
  print("No reciprocal for 0.")
}
```

*The general form:*

    if(boolean_condition{
     expressions
    }else{
     expressions
    })

*An alternative:*

    ifelse()
      a <- 0
    ifelse(a>0, 1/a, NA)
  
Some useful functions:

* any() takes a logical and returns true if any are true
* all() takes a logical and returns true if all are true

*Basic Functions*

Here's an example of a simple function:

```{r}
avg <- function(x){
  s<- sum(x)
  n <- length(x)
  s/n
}
```


Note: variables defined inside a function are not defined outside, so
    + I can define `s` within `avg` and then define `s <- 3` outside avg without           conflict

Here's the general form:

    my_function <- function(x){
      operations_on_x
     value_of_final_line_is_returned
    }

With arguments:

    my_function <- function(x, arithmetic = TRUE){
     n <-length(x)
     ifelse(arithmetic = TRUE, sum(x)/n, prod(x)^(1/n))
    }
  Note: returns arithmetic mean if TRUE and geometric if false

*Using For Loops*

An example:
 
```{r}
compute_s_n <- function(n){
    x <- 1:n
    sum(x)
}
  compute_s_n(4)
```


*General Form of Loops*

    for (i in range_of_values) {
      operations_that_use_i_which_is_changing_across_values
    }

One example:

```{r}
m <- 25
s_n <- vector(length = m) #create an empty vector (necessary for a for loop)
for(n in 1:m){ #create for loop to find product for 1:25
  s_n[n] <- compute_s_n(n)
}
n <- 1:25
plot(n, s_n)
```

*Conditionals*

More function examples:

```{r}
sum_n <- function(n){ #creates a function called sum(n) that creates a series from 1 to         the input, and returns the sum
  x <- 1:n
  y <- sum(x)
  y
}
```

